---
title: "Including demographic processes in a population model"
author: "Jian Yen"
date: "4/11/2020"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(aae.pop)
```

## Demographic processes?

This is a broad term that could mean many things. `aae.pop` focuses on four main processes:

- demographic stochasticity: random variation in the outcomes of individual processes (e.g. birth, death), resulting in variation in population abundances.

- environmental stochasticity: random variation in external conditions, resulting in variation in the vital rates (i.e., the population matrix).

- density dependence: variation in vital rates or population outcomes that depends on the abundance of the population. Density dependence can be negative (e.g., due to limited resources) or positive (e.g., Allee effects).

- covariate effects: the influence of covariates on the population matrix. In `aae.pop`, covariate effects are included primarily to allow the population matrix to change through time in response to external factors (e.g., weather, habitat availability).

Two additional processes, *dispersal* and *interspecific interactions*, are covered in the [Metapopulations](metapopulations.html) and [Multiple species](multiple_species.html) vignettes.

## An aside: `mask`s

The processes above are unlikely to have consistent effects across all vital rates or all classes in a population. To deal with this, `aae.pop` is built around a concept of *masks*. Masks are used to select specific elements of the population matrix, and only these elements are altered by a given process.

Masks are paired with functions when defining all of the above processes. Masks tell R which cells to target, functions tell R what to do to these cells.

Several helper functions are included to define common masks. These include the `reproduction`, `survival`, and `transition` regions of the matrix discussed in the [getting started](get_started.html) vignette, as well as masks to select an entire population matrix (`all_cells`) or abundance vector (`all_classes`).

The `mask`s defined in `aae.pop` return a `TRUE`/`FALSE` matrix or vector that selects the required cells. Masks are defined in this way because `aae.pop` flattens the population matrix in all internal calculations, which breaks cell-based subsetting (i.e., the `[i, j]` R notation will not work). It is possible (but not advised) to work around this. The discussion [below](#nomask) covers how to work around masks and why this might not be a good idea.

## Back to demographic processes

#### A basic model

Let's start with the basic model used in the [getting started](get_started.html) vignette. This was a Leslie matrix with five age classes, with individuals reproducing from ages 3-5.

```{r}
popmat <- rbind(
  c(0,    0,    2,    4,    7),  # reproduction from 3-5 year olds
  c(0.25, 0,    0,    0,    0),  # survival from age 1 to 2
  c(0,    0.45, 0,    0,    0),  # survival from age 2 to 3
  c(0,    0,    0.70, 0,    0),  # survival from age 3 to 4
  c(0,    0,    0,    0.85, 0)   # survival from age 4 to 5
)
```

This is sufficient to start simulating population dynamics, using the `dynamics` and `simulate` functions. However, adding additional processes requires a few more steps first.

#### Demographic stochasticity

Random variation in individual outcomes is a key feature of most population models. This variation is most important (and influential) when populations are small, so that a few small events (e.g., several individuals failing to reproduce) can have a big effect on population outcomes.

Demographic stochasticity will be defined here with a single `mask` and its corresponding `function`. The simplest form of demographic stochasticity in this case is Poisson variation around the expected number of individuals in the next generation. This can be coded up as:

```{r}
demostoch_mask <- all_classes(popmat) # affects all classes
demostoch_fn <- function(x) {
  rpois(length(x), lambda = x)
}
```

Here, the `mask` selects all classes in the population vector and the `function` takes this vector `x` and returns random Poisson variates with mean equal to `x`.

The `mask`/`function` pair can be combined into a single object with the `demographic_stochasticity` function. 

```{r}
demostoch <- demographic_stochasticity(
  masks = demostoch_mask,
  funs = demostoch_fn
)
```

The resulting `demostoch` object can be passed directly to `dynamics`, along with the population matrix. In turn, this creates a `dynamics` object that can be used in `simulate`.

```{r, dev = "png"}
# create population dynamics object
popdyn <- dynamics(popmat, demostoch)

# simulate population dynamics
sims <- simulate(popdyn, nsim = 100)

# plot the population trajectories
plot(sims, col = scales::alpha("#2171B5", 0.1))
```

#### Environmental Stochasticity

Environmental stochasticity will be defined here with two `mask`s and their corresponding `function`s. The simplest form of variation in reproduction is Poisson variation around the expected number of new individuals. This can be coded up as:

```{r}
reproduction_mask <- reproduction(popmat, dims = 3:5) # only ages 3-5 reproduce
reproduction_fn <- function(x) {
  rpois(length(x), lambda = x)
}
```

Here, the `mask` selects the reproductive output of age classes 3 to 5 in the population matrix and the `function` takes this vector `x` and returns random Poisson variates with mean equal to `x`.

The second `mask`/`function` pair will add variation in survival outcomes. In this case, a simple form of variation is to add or subtract some small value from the survival probabilities. This can be coded up as:

```{r}
transition_mask <- transition(popmat) # all classes this time
transition_fn <- function(x) {
  
  # add a random deviation to x
  deviation <- runif(length(x), min = -0.1, max = 0.1)
  x <- x + deviation
  
  # make sure the result isn't negative or greater than 1
  x[x < 0] <- 0
  x[x > 1] <- 1
  
  # return the value
  x
  
}
```

Here, the `mask` selects all elements on the sub-diagonal of the population matrix, and the `function` takes some vector `x` and returns a new value with a deviation between -0.1 and 0.1. There is one extra check in the function to make sure that the new probabilities are still probabilities (i.e., still between 0 and 1).

These two `mask`/`function` pairs can be combined into a single object with the `environmental_stochasticity` function. 

```{r}
envstoch <- environmental_stochasticity(
  masks = list(reproduction_mask, transition_mask),
  funs = list(reproduction_fn, transition_fn)
)
```

The resulting `envstoch` object can be passed directly to `dynamics`, along with the population matrix. In turn, this creates a `dynamics` object that can be used in `simulate`.

```{r, dev = "png"}
# create population dynamics object
popdyn <- dynamics(popmat, envstoch)

# simulate population dynamics
sims <- simulate(popdyn, nsim = 100)

# plot the population trajectories
plot(sims, col = scales::alpha("#2171B5", 0.1))
```

It is easy to include both environmental and demographic stochasticity in the same model. The `dynamics` call simply needs both `envstoch` and `demostoch` objects:

```{r, dev = "png"}
# create population dynamics object
popdyn <- dynamics(popmat, envstoch, demostoch)

# simulate population dynamics
sims <- simulate(popdyn, nsim = 100)

# plot the population trajectories
plot(sims, col = scales::alpha("#2171B5", 0.1))
```

#### Density dependence

Density dependence is the effects of population abundances on vital rates or individual outcomes. The most common form of density dependence in population ecology is *negative density dependence*, that is, a reduction in vital rates with increases in population abundance. Negative density dependence is typically reflected in changes in reproduction due to, for example, increased competition for resources resulting in high mortality of new individuals.

Density dependence can be included in exactly the same way as environmental stochasticity: a `mask` tells R which cells in the population matrix are affected by abundances and a `function` tells R what to do with these cells. In this case, the `function` can see the vital rates but also the abundances in each population class.

```{r}
reproduction_mask <- reproduction(popmat, dims = 3:5) # only ages 3-5 reproduce
reproduction_fn <- function(x) {
  rpois(length(x), lambda = x)
}
```

There are other forms of density dependence. Positive. And negative can affect adults etc. All use same approach (but see following section for exception).

#### A different form of density dependence

ADD

#### Covariates

ADD

## What if I don't want to use masks? {#nomask}

Why this is hard (and a bad idea unless literally operating on the whole matrix at once).

Workaround for passing dims to functions.
