---
title: "Including demographic processes in a population model"
author: "Jian Yen"
date: "5/11/2020"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(aae.pop)
```

## Demographic processes?

This is a broad term that could mean many things. `aae.pop` focuses on four main processes:

- demographic stochasticity: random variation in the outcomes of individual processes (e.g. birth, death), resulting in variation in population abundances.

- environmental stochasticity: random variation in external conditions, resulting in variation in the vital rates (i.e., the population matrix).

- density dependence: variation in vital rates or population outcomes that depends on the abundance of the population. Density dependence can be negative (e.g., due to limited resources) or positive (e.g., Allee effects).

- covariate effects: the influence of covariates on the population matrix. In `aae.pop`, covariate effects are included primarily to allow the population matrix to change through time in response to external factors (e.g., weather, habitat availability).

Two additional processes, *dispersal* and *interspecific interactions*, are covered in the [Metapopulations](metapopulations.html) and [Multiple species](multiple_species.html) vignettes.

## An aside: masks

The processes above are unlikely to have consistent effects across all vital rates or all classes in a population. To deal with this, `aae.pop` is built around a concept of *masks*. Masks are used to select specific elements of the population matrix, and only these elements are altered by a given process.

Masks are paired with functions when defining all of the above processes. Masks tell R which cells to target, functions tell R what to do to these cells.

Several helper functions are included to define common masks. These include the `reproduction`, `survival`, and `transition` regions of the matrix discussed in the [getting started](get_started.html) vignette, as well as masks to select an entire population matrix (`all_cells`) or abundance vector (`all_classes`).

The masks defined in `aae.pop` return a `TRUE`/`FALSE` matrix or vector that selects the required cells. Masks are defined in this way because `aae.pop` flattens the population matrix in all internal calculations, which breaks cell-based subsetting (i.e., the `[i, j]` R notation will not work). It is possible (but not advised) to work around this. The discussion [below](#nomask) introduces one way to work around masks and highlights why this might not be a good idea.

## Back to demographic processes

### A basic model

Let's start with the basic model used in the [Getting started](get_started.html) vignette. This was a Leslie matrix with five age classes, with individuals reproducing from ages 3-5.

```{r}
popmat <- rbind(
  c(0,    0,    2,    4,    7),  # reproduction from 3-5 year olds
  c(0.25, 0,    0,    0,    0),  # survival from age 1 to 2
  c(0,    0.45, 0,    0,    0),  # survival from age 2 to 3
  c(0,    0,    0.70, 0,    0),  # survival from age 3 to 4
  c(0,    0,    0,    0.85, 0)   # survival from age 4 to 5
)
```

This is sufficient to start simulating population dynamics, using the `dynamics` and `simulate` functions. However, adding additional processes requires a few more steps first.

### Demographic stochasticity

Random variation in individual outcomes is a key feature of most population models. This variation is most important (and influential) when populations are small, so that a few small events (e.g., several individuals failing to reproduce) can have a big effect on population outcomes.

Demographic stochasticity will be defined here with a single `mask` and its corresponding `function`. The simplest form of demographic stochasticity in this case is Poisson variation around the expected number of individuals in the next generation. This can be coded as:

```{r}
demostoch_mask <- all_classes(popmat) # affects all classes
demostoch_fn <- function(x) {
  rpois(length(x), lambda = x)
}
```

Here, the `mask` selects all classes in the population vector and the `function` takes this vector `x` and returns random Poisson variates with mean equal to `x`. Note that this function potentially allows the number of surviving individuals in a class to exceed the number of available individuals because the Poisson distribution does not have an upper bound. A workaround for this, using the Binomial distribution, is covered in the [Beyond defaults](beyond_defaults.html) vignette.

The `mask`/`function` pair can be combined into a single object with the `demographic_stochasticity` function. 

```{r}
demostoch <- demographic_stochasticity(
  masks = demostoch_mask,
  funs = demostoch_fn
)
```

The resulting `demostoch` object can be passed directly to `dynamics`, along with the population matrix. In turn, this creates a `dynamics` object that can be used in `simulate`.

```{r, dev = "png"}
# create population dynamics object
popdyn <- dynamics(popmat, demostoch)

# simulate population dynamics
sims <- simulate(popdyn, nsim = 100)

# plot the population trajectories
plot(sims, col = scales::alpha("#2171B5", 0.1))
```

### Environmental Stochasticity

Environmental stochasticity will be defined here with two masks and their corresponding functions. The simplest form of variation in reproduction is Poisson variation around the expected number of new individuals. This can be coded up as:

```{r}
reproduction_mask <- reproduction(popmat, dims = 3:5) # only ages 3-5 reproduce
reproduction_fn <- function(x) {
  rpois(length(x), lambda = x)
}
```

Here, the `mask` selects the reproductive output of age classes 3 to 5 in the population matrix and the `function` takes this vector `x` and returns random Poisson variates with mean equal to `x`.

The second `mask`/`function` pair will add variation in survival outcomes. In this case, a simple form of variation is to add or subtract some small value from the survival probabilities. This can be coded up as:

```{r}
transition_mask <- transition(popmat) # all classes this time
transition_fn <- function(x) {
  
  # add a random deviation to x
  deviation <- runif(length(x), min = -0.1, max = 0.1)
  x <- x + deviation
  
  # make sure the result isn't negative or greater than 1
  x[x < 0] <- 0
  x[x > 1] <- 1
  
  # return the value
  x
  
}
```

Here, the `mask` selects all elements on the sub-diagonal of the population matrix, and the `function` takes some vector `x` and returns a new value with a deviation between -0.1 and 0.1. There is one extra check in the function to make sure that the new probabilities are still probabilities (i.e., still between 0 and 1).

These two `mask`/`function` pairs can be combined into a single object with the `environmental_stochasticity` function. 

```{r}
envstoch <- environmental_stochasticity(
  masks = list(reproduction_mask, transition_mask),
  funs = list(reproduction_fn, transition_fn)
)
```

The resulting `envstoch` object can be passed directly to `dynamics`, along with the population matrix. In turn, this creates a `dynamics` object that can be used in `simulate`.

```{r, dev = "png"}
# create population dynamics object
popdyn <- dynamics(popmat, envstoch)

# simulate population dynamics
sims <- simulate(popdyn, nsim = 100)

# plot the population trajectories
plot(sims, col = scales::alpha("#2171B5", 0.1))
```

It is easy to include both environmental and demographic stochasticity in the same model. The `dynamics` call simply needs both `envstoch` and `demostoch` objects:

```{r, dev = "png"}
# create population dynamics object
popdyn <- dynamics(popmat, envstoch, demostoch)

# simulate population dynamics
sims <- simulate(popdyn, nsim = 100)

# plot the population trajectories
plot(sims, col = scales::alpha("#2171B5", 0.1))
```

### Density dependence

Density dependence occurs when vital rates or individual outcomes depend on population abundance. The most common form of density dependence in population ecology is *negative density dependence*, that is, a reduction in vital rates with increases in population abundance. Negative density dependence is typically observed as changes in reproduction due to, for example, increased competition for resources resulting in high mortality of new individuals.

Density dependence can be included in exactly the same way as environmental stochasticity: a `mask` tells R which cells in the population matrix are affected by abundances and a `function` tells R what to do with these cells. In this case, the `function` is passed the vital rates as well as the vector of abundances in each population class.

`aae.pop` has pre-defined functions for two common forms of density dependence: the *Ricker* model and the *Beverton-Holt* model. The Ricker model assumes that mortality rates of new individuals are proportional to adult population size due to *scramble competition* (equal division of resources). The Beverton-Holt model assumes that mortality rates of new individuals are linearly dependent on the number of individuals due to *contest competition* (unequal division of resources). Importantly, the Ricker model is *overcompensatory*, which means it can increase mortality rates to the point of complete mortality. This does not occur under the Beverton-Holt model.

In `aae.pop`, both models are specified with one parameter, `k`, that represents the *carrying capacity* of the population. Both functions include one extra argument, `exclude`, which allows estimates of carrying capacity to be defined for a subset of population classes (e.g., adult abundances only). A Ricker model can be specified as follows:

```{r, dev = "png"}
# specify a Ricker model for density dependence
dd <- density_dependence(
  masks = reproduction(popmat, dims = 3:5), # only adults reproduce
  funs = ricker(k = 40, exclude = 1:2)      # set k based on adult abundances
)

# create population dynamics object
popdyn <- dynamics(popmat, envstoch, demostoch, dd)

# simulate population dynamics
sims <- simulate(popdyn, nsim = 100)

# plot the population trajectories
plot(sims, col = scales::alpha("#2171B5", 0.1))
```

Density dependence can take many other forms. These forms may include positive density dependence (e.g., Allee effects) or may affect adult abundances as well as reproduction. An example of the former is included in the [Macquarie perch worked example](macperch_example.html). The next section introduces an example of the latter.

### A different form of density dependence

Density dependence is most commonly assumed to affect only the new individuals entering the population. However, it is possible that density dependence also affects adults, for example, when high abundances limit access to suitable habitat. Effects on adult survival could be modelled using the approach outlined in the previous section, with `transition` or `survival` masks alongside the `reproduction` mask.

An alternative approach is to model density dependence through its direct effects on population abundances. Although not linked directly to vital rates (e.g., survival), this approach can be easier to implement and can capture processes that do not directly affect vital rates, such as fishing, hunting, or logging.

The `density_dependence_n` function defines this form of density dependence. In this case, the `mask` needs to select the relevant classes in the population, and the `function` takes only one argument, the vector of population abundances. This approach can be implemented as follows:

```{r, dev = "png"}
# specify density dependence that removes 10 % of adults in each generation
dd_n <- density_dependence_n(
  masks = all_classes(popmat, dims = 3:5), # want to focus on adults
  funs = function(n) 0.9 * n               # define in-line function to return
                                           #   90 % of adult population
)

# create population dynamics object
popdyn <- dynamics(popmat, envstoch, demostoch, dd_n)

# simulate population dynamics
sims <- simulate(popdyn, nsim = 100)

# plot the population trajectories
plot(sims, col = scales::alpha("#2171B5", 0.1))
```

### Covariates

Population dynamics often depend on external factors (`covariates`, here). `aae.pop` uses the `mask`/`function` approach described above to include the effects of covariates on vital rates. The `function` takes the vital rates as the first argument, followed by any arguments required to specify covariate values or their effects. An example illustrates this most clearly:

```{r, dev = "png"}
# set up 20 years of covariates, such as proportion of available nesting sites
nesting <- runif(20, min = 0.5, max = 1.0)    # 50 % to 100 % of total sites available in each year

# define the mask
covar_mask <- reproduction(popmat, dims = 3:5)  # assume nesting sites only affect reproduction

# define a function that links vital rates to the covariates
covar_fn <- function(x, nests) {
  x * nests    # really simple function, assume the proportion of successful reproduction
               #   attempts is equal to the proportion of nests
}

# combine this into a covariates term
covs <- covariates(covar_mask, covar_fn)

# create population dynamics object (without density dependence)
popdyn <- dynamics(popmat, envstoch, demostoch, covs)

# simulate population dynamics
sims <- simulate(
  popdyn,
  nsim = 100,
  args = list(covariates = format_covariates(nesting))
)

# plot the population trajectories
plot(sims, col = scales::alpha("#2171B5", 0.1))
```

Specifying a `covariates` term does not automatically include covariates in the model. Note that the `nesting` variable also had to be passed to `simulate` via the `args` argument. The reason for this is that it allows a single population dynamics object (`popdyn`, above) to be used with multiple different sets of covariates, without re-compiling the `dynamics` object each time. The `format_covariates` function is included to format covariate values and any auxiliary (static) variables.

`aae.pop` assumess that `covariates` change through time, so a consequence of specifying covariates is that the vector/matrix of covariates determines the number of simulated time steps (`n_time`). In the above example, there were 20 values of `nesting`, which translates to 21 time steps (initial condition plus 20 updates).

It is assumed that covariates can be passed as a vector, matrix, or data.frame, with one element or row for each time step. More complex `covariate` terms are possible through the `args.dyn` and `args.fn` arguments to `simulate`. These terms are demonstrated in the [Passing arguments](passing_arguments.html) vignette.

## What if I don't want to use masks? {#nomask}

The `mask`/`function` is central to `aae.pop`. This approach allows multiple functions to be combined into a single term and makes it easier to keep track of (and update) individual functions or masks. A secondary reason for this approach is computational. Rarely will any demographic processes act on the entire population matrix. When dealing with large matrices (e.g., 50-100 classes), it's much more efficient to pass small parts of the matrix than the entire matrix.

There might be situations where the entire population matrix does need to be modified by a given process. This is still supported in `aae.pop` through the `all_cells` `mask`. However, it isn't possible to subset the matrix within a `function` using R's standard `[i, j]` notation. This is because the functions within `aae.pop` receive a flattened (and masked) version of the population matrix, which actually looks like:

```{r, echo = FALSE}
# reproduction mask
popmat[reproduction(popmat, dims = 3:5)]
```

or 

```{r, echo = FALSE}
# transition mask
popmat[transition(popmat)]
```

or

```{r, echo = FALSE}
# all elements
popmat[all_cells(popmat)]
```

If it's essential to use `[i, j]` subsetting inside a function, a workaround is to reconstruct the matrix within the function by passing the dimensions through the `args` argument in `simulate`. This might look a bit like:

```{r, dev = "png"}
# define a function
no_mask_fn <- function(x, dim) {
  
  # reconstruct matrix
  x <- array(x, dim = dim)
  
  # change elements with [i, j] notation
  x[2, 1] <- 0.6 * x[2, 1]
  x[3, 2] <- 1.25 * x[3, 2]
  
  # return
  x
  
}

# set this up as a form of environmental stochasticity
envstoch <- environmental_stochasticity(
  masks = all_cells(popmat),  # pass the entire matrix
  funs = no_mask_fn           # use the no-mask function
)

# create population dynamics object (without density dependence)
popdyn <- dynamics(popmat, envstoch)

# simulate population dynamics
sims <- simulate(
  popdyn,
  nsim = 100,
  args = list(environmental_stochasticity = list(dim = dim(popmat)))
)

# plot the population trajectories
plot(sims, col = scales::alpha("#2171B5", 0.1))
```
