---
title: "Population model templates with aae.pop.templates"
author: "Jian Yen"
date: "01/07/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(aae.pop)
library(aae.pop.templates)
```

## Background

A major benefit of `aae.pop` is that a single population dynamics object can be used in multiple applications, each with different parameters, model settings, or processes. To support this, the `aae.pop.templates` package has been developed to collate pre-defined population models for several species. The `aae.pop.templates` package is separated from the `aae.pop` package because it is limited in its current applications and is updated much more regularly than `aae.pop`. This package is not required to use `aae.pop` and is described here primarily to illustrate how templates can support applications where a single model structure is used frequently. Current templates focus primarily on models of freshwater fish species. All models were developed by researchers at the Arthur Rylah Institute for Environmental Research. 

This document lists several templates included in the `aae.pop.templates` package, outlines the structure of a basic template, and demonstrates how to use a template.

## Installing `aae.pop.templates`

You can install the `aae.pop.templates` package from GitHub. To install from GitHub, you'll need the `remotes` R package, which you can install from CRAN with:

```{r, eval = FALSE}
install.packages("remotes")
```

You can then install the `aae.pop.templates` package from Github with:

```{r, eval = FALSE}
remotes::install_github("aae-stats/aae.pop.templates")
```

Once completed, you should be able to load the 
`aae.pop.templates` package with `library(aae.pop.templates)`.

## What is a template?

The purpose of a template is to capture the population dynamics object for a species, along with additional arguments or functions required to simulate population dynamics. This is slightly more involved than simply defining the population matrix and some default processes. Rather, a template should be flexible enough to encompass multiple scenarios or contexts and should be supported by a clear rationale for model structure and parameters. In many cases, this will involve references to published works detailing a model's development and implementation.

Templates have been developed for several freshwater fish species, including Murray cod (*Maccullochella peelii*), Macquarie perch (*Macquaria australasica*), estuary perch (*Percalates colonorum*), Yarra pygmy perch (*Nannoperca obscura*), barred galaxias (*Galaxias fuscus*), and Australian bass (*Percalates novemaculeata*). An additional template has been developed for platypus (*Ornithorhynchus anatinus*).

## Template structure

Templates have their own object class. This class is handled slightly differently to standard `dynamics` objects in `aae.pop`. This structure is required because templates comprise a `dynamics` object as well as a set of arguments that specify how a simulation should be conducted. Templates capture quite complex models, so it is common that the simulation process for these models depends on non-standard parameters or functions.

In `aae.pop.templates`, templates are defined by a function following the standard naming structure `template_SPECIES`, where `SPECIES` is the name of the relevant species. This function can take any arguments and needs to return a named list with an element named `dynamics` containing the components of a `dynamics` object (not a compiled `dynamics` object) and optionally an element named `arguments` containing a compiled set of arguments to be passed to `simulate` via the `args` argument. Arguments are compiled with the `get_args` function, which calls a function with the naming structure `args_SPECIES`, where `SPECIES` is the name of the relevant species. An `args_SPECIES` function can take any arguments and needs to return a named list as would be passed to `args` in `simulate` (e.g., arguments specifying the form of `density_dependence` or `covariates`). Finally, templates in `aae.pop.templates` are wrapped in a single function that calls `get_template`, which compiles the population `dynamics` object and returns a `template` object.

Although relatively complex, a straightforward way to develop a template is to copy an existing template and update specific details as required. Including a custom template in `aae.pop.templates` requires several more steps because both `get_template` and `get_args` check that a template exists for a given species with the `aae.pop.templates:::check_species_template` and `aae.pop.templates:::check_species_args` functions. The general steps outlined above would work similarly for a custom template not included in `aae.pop.templates` but would require custom functions to compile arguments and `dynamics` objects because `get_args` and `get_template` will throw an error.

## Using templates

Once defined, templates are straightforward to use. Here is an example simulating Murray cod dynamics for 50 years:

```{r, dev = "png"}
mc <- murray_cod()
sims <- simulate(mc, nsim = 100)
plot(sims, col = "#2171B5")
```

Note that, with three lines of code, this template now simulates 100 replicate trajectories for a species with 50 age classes and complex forms of density dependence and environmental and demographic stochasticity. Importantly, this template also incorporates river-specific effects of covariates, stocking and angling effects, and variable carrying capacities. Changing these requires either a few arguments to the template definition or, in the case of covariates, requires covariates provided as `args` to `simulate`.

```{r, dev = "png"}
# change carrying capacity (applies to adults, total plotted
#   abundances will exceed this value)
mc <- murray_cod(k = 1000)
sims <- simulate(mc, nsim = 100)
plot(sims, col = "#2171B5", main = "K = 1000")

# add angling effects
mc <- murray_cod(k = 10000, p_capture = 0.2)
sims <- simulate(mc, nsim = 100)
plot(sims, col = "#2171B5", main = "20 % angler catch rate")

# add angling effects
mc <- murray_cod(k = 10000, p_capture = 0.1, slot = c(500, 800))
sims <- simulate(mc, nsim = 100)
plot(sims, col = "#2171B5", main = "10 % angler catch rate but relaxed restrictions")

# add angling effects
mc <- murray_cod(k = 10000, p_capture = 0.1, n = c(1000, 0, 0))
sims <- simulate(mc, nsim = 100)
plot(sims, col = "#2171B5", main = "10 % angler catch rate and 1000 fingerlings stocked annually")
```

Arguments to `murray_cod` are described in the associated help file (`?murray_cod`), as are those for all other species templates. Covariates are not currently described in these documents but some features to interrogate and specify covariates are in development.
